Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-03-18 paperDistance/paper.tex
subsection heading
references

The \gaia\ Mission (DWH CITE) will deliver more than a billion stellar distances.
Only a small fraction (but large number) of these distance
measurements will be purely astrometric:
\gaia\ uses astrometric parallax to determine the distances of the closer
stars, and calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $B_p\,R_p$ spectrophotometry, are used to provide
distance estimates to more distant stars (DWH CITE).
The full stack required for these distance inferences is complex.
It involves modeling not just stars, but also the dust in the Milky Way,
and the response of the telescope itself (DWH CITE).

With projects like \project{The Cannon} (DWH CITE) and \project{Avast}
(DWH CITE), we are exploring the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate information than any theoretical or
physics-based model?
The answer to this question is extremely context-dependent: It depends
on what data are available, how much data are available, and what
questions are being asked.
In general, data-driven models contain fewer assumptions than
physics-driven models, but they are also usually less interpretable.
However, the \gaia\ use case is ideal for thinking about purely statistical
models of stars:
The distance to a star is a simple latent variable, and the nearby
stars constitute a high signal-to-noise ``training set''.
These nearby stars can be used to anchor empirical models that can
then be applied to the more distant stars.

@@

Here we build and use just such a data-driven model.  In particular,
we use all of the \gaia\ \tgas\ data, matched to the
\tmass\ photometry (\citealt{skrutskie06}), and in the footprint of
the \panstarrs\ Dust Map (CITE GREEN), to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We use an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (DWH CITE BOTH) and to perform quasar target selection
in \project{Sloan Digital Sky Survey} data (DWH CITE BOTH).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its principle disadvantage for the present purposes is that it requires
or implies a strictly Gaussian noise model.
In the below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

@@

It wouldn't be using the data twice---that is, it would be valid from a
probabilistic perspective---if we performed a full hierarchical Bayesian
inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to a related
project (CITE BORIS), in which we use a slightly less appropriate
model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

DWH: Placeholder for the output format for our data products. Should
be the last paragraph or two of this section.

@@

The Empirical Bayesian methodology is a good approximation to full
hierarchical inference in the limit of large numbers of data points.
That is, when no individual star is carrying a lot of weight in the
model of the \cmd.
With its millions of stars, the \tgas\ Catalog may appear to safely
live in this limit.
However, because there are parts of the \cmd\ that are poorly populated,
even in \tgas, the Empirical Bayes approximation may be bad for some
color ranges.
In particular, there appear to be essentially no white dwarf stars,
and other classes (like hot main-sequence stars) have few to no exemplars
in the data set at high signal-to-noise (in parallax).
For all these reasons, we do think that this approximation is not perfectly
safe, and it is a goal to explore full hierarchical modeling in the coming
years (and see, for example, CITE BORIS).

@@

The outputs of this project include posterior pdfs and posterior
samplings for stellar parallaxes or distances.
It is tempting to treat these outputs as equivalent to some kind of catalog of
measurements, as we treated the original parallax measurements from \tgas.
That is, it is tempting to treat these as simply ``better'' measurements.
Taken one star at a time, this is permitted and true, and the basis for our
claim that we have de-noised the \tgas\ Catalog.
However, there are important differences, as there are with all
probabilistic catalogs (HOGG CITE LANG, FINKBEINER), between the
measurements and the de-noised measurements.
The most important difference, for users of our output, between our
output and the \tgas\ data input, is that the \tgas\ data is the representation
of a likelihood function, whereas our ouptut is a
representation of a posterior pdf, one per star.
If used naively, posterior information can lead to serious statistical
errors, because each data point has had a prior multiplied in.
Effectively multiplicative uses of the data will effectively take the prior
pdf to a large power.
We have (in previous work)
given examples of correct uses of posterior samplings or pdfs
for subsequent inferences (\citealt{hogg08}; \citealt{dfm14}); we
encourage power-users to consult those methodological
contributions before engaging with these outputs.

>>>>

The \gaia\ Mission (\citealt{gaia16}) will deliver more than a billion stellar distances.
Only a small fraction (but large number) of these distance
measurements will be purely astrometric:
\gaia\ uses astrometric parallax to determine the distances of the closer
stars, and calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $B_p\,R_p$ spectrophotometry, are used to provide
distance estimates to more distant stars.
The full stack required for these distance inferences is complex.
It involves modeling not just stars, but also the dust in the Milky Way,
and the response of the telescope itself.

With projects like \project{The Cannon} (\citealt{ness15}; \citealt{casey17}; \citealt{ho17})
and \project{Avast} (Bedell et al.\ in preparation),
we are exploring the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate information than any theoretical or
physics-based model?
The answer to this question is extremely context-dependent: It depends
on what data are available, how much data are available, and what
questions are being asked.
In general, data-driven models contain fewer assumptions than
physics-driven models, but they are also usually less interpretable.
However, the \gaia\ use case is ideal for thinking about purely statistical
models of stars:
The distance to a star is a simple latent variable, and the nearby
stars constitute a high signal-to-noise ``training set''.
These nearby stars can be used to anchor empirical models that can
then be applied to the more distant stars.

@@

Here we build and use just such a data-driven model.  In particular,
we use all of the \gaia\ \tgas\ data (\citealt{tgas}), matched to the
\tmass\ photometry (\citealt{skrutskie06}), and in the footprint of
the \panstarrs\ Dust Map (CITE GREEN), to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We use an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (\citealt{hogg05}; \citealt{bovy09}) and to
perform quasar target selection
in \project{Sloan Digital Sky Survey} data (\citealt{xdqso}; \citealt{xdqsoz}).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its principle disadvantage for the present purposes is that it requires
or implies a strictly Gaussian noise model.
In the below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

@@

It wouldn't be using the data twice---that is, it would be valid from a
probabilistic perspective---if we performed a full hierarchical Bayesian
inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to a related
project (Leistedt et al. in preparation), in which we use a slightly less appropriate
model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

\subsection{Output data}
DWH: Placeholder for the output format for our data products.

@@

The Empirical Bayesian methodology is a good approximation to full
hierarchical inference in the limit of large numbers of data points.
That is, when no individual star is carrying a lot of weight in the
model of the \cmd.
With its millions of stars, the \tgas\ Catalog may appear to safely
live in this limit.
However, because there are parts of the \cmd\ that are poorly populated,
even in \tgas, the Empirical Bayes approximation may be bad for some
color ranges.
In particular, there appear to be essentially no white dwarf stars,
and other classes (like hot main-sequence stars) have few to no exemplars
in the data set at high signal-to-noise (in parallax).
For all these reasons, we do think that this approximation is not perfectly
safe, and it is a goal to explore full hierarchical modeling in the coming
years (and see, for example, Leistedt et al. in preparation).

@@

The outputs of this project include posterior pdfs and posterior
samplings for stellar parallaxes or distances.
It is tempting to treat these outputs as equivalent to some kind of catalog of
measurements, as we treated the original parallax measurements from \tgas.
That is, it is tempting to treat these as simply ``better'' measurements.
Taken one star at a time, this is permitted and true, and the basis for our
claim that we have de-noised the \tgas\ Catalog.
However, there are important differences, as there are with all
probabilistic catalogs (\citealt{hogg11}; \citealt{portillo17}), between the
measurements and the de-noised measurements.
The most important difference, for users of our output, between our
output and the \tgas\ data input, is that the \tgas\ data is the representation
of a likelihood function, whereas our ouptut is a
representation of a posterior pdf, one per star.
If used naively, posterior information can lead to serious statistical
errors, because each data point has had a prior multiplied in.
Effectively multiplicative uses of the data will effectively take the prior
pdf to a large power.
We have (in previous work)
given examples of correct uses of posterior samplings or pdfs
for subsequent inferences (\citealt{hogg08}; \citealt{dfm14}); we
encourage power-users to consult those methodological
contributions before engaging with these outputs.
